---
title: "basic syntax"
output: 
  
  html_notebook:
    theme: united
    toc: yes
  html_document:
    toc: yes
---

Note: to run the lines, select them or locate the cursor on a specific line and hold the `ctrl` key while typing `enter`  

### basic math operations    

We can perform calculations on numerical objects
```{r}
1 + 1
```

```{r}
1 - 1
```

```{r}
1 / 1
```

```{r}
1 * 1
```

### classes  

To find out what class an object is you can use the `class()` function  
```{r}
class(1)
```

Another class is the so called `string` or `character` type,  
you let R know you want to define a character class by using single or double quotes:  
```{r}
class("1")
```

* exercise: Run the line, why did you get an error?   
Can you fix this error? 
```{r}
13 + "5"
```

The `+` is an example of a binary operator. Basically a function that can take 2 input arguments and return a single output.  
However, we will look into functions later today. The `non-numeric` argument refers to the 5 not being numeric. It is called an `argument` because it is the input to the binary operator.  

### objects for storing data  
There are several objects in R that handle data storage. We will go through them one by one.   
But first the concept of placeholders, also called identifiers, has to be explained.  

You, as a programmer, are expected to create logical placeholder names. These names should be short, but still clarify what it is. Placeholders are used to store any object type. It is recommended to:  
* seperate words with underscores: my_variable.  
* use small letters  
* for functions use verbs: join_tables()  

As an example of a placeholder, let us define a variable that enables us to store a time-stamp:  
```{r}
current_time <- Sys.time()
current_time  
```
Run the lines several times, see how the placeholder `current_time` stores the output of our `Sys.time()` function?  
We can also store the output of a calculation:  

```{r}
my_result <- 5 * (1+5)
my_result
```
And use this later on in our analysis, becuase it is stored in `R memory`! 
```{r}
my_result + 1
```

* exercise: explain and then fix the error message you get when running the following line of code: 
```{r}
my_other_result + 1  
```




#### vector
A vector object is created with the `c()` command, each entry is seperated by a comma.  

```{r}

```

### accessing and selection   

