---
title: "basic syntax"
output: 
  
  html_notebook:
    theme: united
    toc: yes
  html_document:
    toc: yes
editor_options: 
  chunk_output_type: inline
---

Note: to run the lines, select them or locate the cursor on a specific line and hold the `ctrl` key while typing `enter`  

## Basic math operations, classes and variables        
### basic math operations    

We can perform calculations on numerical objects
```{r}
1 + 1
```

```{r}
1 - 1
```

```{r}
1 / 1
```

```{r}
1 * 1
```

### classes  

To find out what class an object is you can use the `class()` function  
```{r}
class(1)
```

Another class is the so called `string` or `character` type,  
you let R know you want to define a character class by using single or double quotes:  
```{r}
class("1")
```

* exercise: Run the line, why did you get an error?   
Can you fix this error? 
```{r}
13 + "5"  
```

The `+` is an example of a binary operator. Basically a function that can take 2 input arguments and return a single output.  
However, we will look into functions later today. The `non-numeric` argument refers to the 5 not being numeric. It is called an `argument` because it is the input to the binary operator.  

Another important class in R is the `factor`. Factors allow ordering of descriptors/ nominal variables. Many R functions treat factors differently compared to the character class. It is important to be aware if your data is stored as character vs factor.  
The default in R is to store all character data as factor. You can turn this default behavior off with `options(stringsAsFactors = FALSE)`. When we look at the data object class `data.frames` we will have a look at some examples involving factors.  



## data storage objects  
There are several objects in R that handle data storage. We will go through them one by one.   
But first the concept of placeholders, also called identifiers, has to be explained.  

You, as a programmer, are expected to create logical placeholder names. These names should be short, but still clarify what it is. Placeholders are used to store any object type. It is recommended to:  
* separate words with underscores: my_variable.  
* use small letters  
* for functions use verbs: join_tables()  


### placeholders  
As an example of a placeholder, let us define a variable that enables us to store a time-stamp:  
```{r}
current_time <- Sys.time()
current_time  
```
Run the lines several times, see how the placeholder `current_time` stores the output of our `Sys.time()` function?  
We can also store the output of a calculation:  

```{r}
my_result <- 5 * (1+5)
my_result
```
And use this later on in our analysis, because it is stored in `R memory`! 
```{r}
my_result + 1
```

* exercise: explain and then fix the error message you get when running the following line of code: 
```{r}
my_other_result + 1  
```




### vectors
A vector object is created with the `c()` command, each entry is separated by a comma:  
`my_vector <- c( 4, 6, 2, 4, 1)`   


* exercise: Now that we know how to create vectors, how to create variables and strings/ characters lets practice this:  
Create a vector with the `c()` function containing the sentence, "hello, my name is " in the first entry and in the second vector entry a placeholder containing your name.  Make sure to define the placeholder containing your name first, or you will get an error!   
```{r}

```


Another way to create sequences of numbers is with the `seq()` function. Have a look at the help file for `seq' by typing
?seq  
Note the `from` `to` `by` and `length.out` arguments. To create a sequence of 6 equally spaced numbers from 0 to 10 you can dO:
`my_seq <- seq(from = 0, to = 10, by = 2)` or you can do `my_seq <- seq(from = 0, to = 10, length.out = 6)` 
Another useful method to generate sequences of integers is the colon: `1:10`  

* exercise: Try all three methods to generate a sequence of numbers from 12 to 2 with steps of 1. 
```{r}

```


It is also possible to do calculations with numeric vectors. There is a behavior to be aware of called recycling. 
* exercise: Calculate the sum of these vectors and explain recycling:  `1:6 + 1:2` also try `1:5` + `1:2`, when and why did you get a warning?   
```{r}

```

Now let us move on to selection operations on vectors. Square brackets `[]` are used for selection. Selection by dimension (rows/columns/ or 3d) can be performed by separating the arguments inside the brackets by comma's. A vector however, has only 1 dimension. To select the second entry of my_seq: `my_seq[2]` 
exercise: Select entry 1, 3 and 5 of `my_seq`:  
```{r}

```

### matrixes  
Single class tables can be stored as matrices. Matrices are not so useful for us biologists, but super useful for fast matrix calculations in the for of linear algebra needed for the various statistical tools available in R.  We will have a quick look at them. To define a matrix:  
```{r}
my_matrix <- matrix(1:9, nrow = 3, ncol = 3)
my_matrix
my_matrix2 <- matrix(1:9, nrow = 3, ncol = 3, byrow = TRUE)
my_matrix2
```
Selection of matrix data is two dimensional, you need row and column identifiers, selection occurs like this: 
`my_matrix[`*row-numbers*`,` *column-numbers*`]` to select row 2 and 3 of the 1st and 3rd columns:  

```{r}
my_matrix[2:3, c(1, 3)]  
```


### data.frames  
The data.frame allows mixing data classes, so you can mix numeric, strings and factors. Let's have a look at the already available `iris` data.frame using the `head` function to display only the first 7 rows.   
```{r}
head(iris, n = 7)
```

The column classes are already displayed under the column names (if you have a recent version of R Studio installed). An alternative to inspect the column classes is to check them manually, for column one and five this would be:  
```{r}
class(iris[, 1])
class(iris[, 5])
```
This mixing of data types is not possible in matrices.  
Before we dive deeper into `data.frames`, we will discuss the `factor` class. Let us print the contents of the `Species` column from the `iris` `data.frame`:  
```{r}
iris[, 5]
```
Note the `Levels` at the bottom? This informs us this column has the class `factor`, and also informs us of the unique factor levels. This is mainly important for statistical analysis, where defining the order is important. However many functions and visualization tools in R also use the factor data type to define for example the order of displaying data labels or the order of plotting nominal values on the x-axis.  
You can create the `factor` class with (first we transform the `Species` to character class): 
```{r}
iris$Species <- as.character(iris$Species)
iris$Species
iris$Species <- factor(iris$Species)
iris$Species
```
 Always think of `data.frames` in a column-wise fashion. A column represents a variable. Next to `iris[, 5]` We can aces `data.frame` columns with `iris$Species` or `iris[, "Species"]`  

This is an opportune moment to explain boolean and conditional selection.  
 
 *exercise: Display the `Petal.Length` for all plants that belong to the virginica Species.  
```{r}

```





### lists  


### arrays  


 



## control flow and repetition  

### conditionals

### for loops

### functional programming: apply family  


## functions 


## data reshaping and summarization  


###
###
###



## data visualization  

### base plotting methods  

method dispatch 


### ggplot2 



## read and write  



## packages, CRAN & Bioconductor  

