---
title: "basic syntax"
output: 
  
  html_notebook:
    theme: united
    toc: yes
  html_document:
    toc: yes
editor_options: 
  chunk_output_type: inline
---

Note: to run the lines, select them or locate the cursor on a specific line and hold the `ctrl` key while typing `enter`  

## Basic math operations, classes and variables        
### basic math operations    

We can perform calculations on numerical objects
```{r}
1 + 1
```

```{r}
1 - 1
```

```{r}
1 / 1
```

```{r}
1 * 1
```

### classes  

To find out what class an object is you can use the `class()` function  
```{r}
class(1)
```

Another class is the so called `string` or `character` type,  
you let R know you want to define a character class by using single or double quotes:  
```{r}
class("1")
```

* exercise: Run the line, why did you get an error?   
Can you fix this error? 
```{r}
13 + "5"  
```

The `+` is an example of a binary operator. Basically a function that can take 2 input arguments and return a single output.  
However, we will look into functions later today. The `non-numeric` argument refers to the 5 not being numeric. It is called an `argument` because it is the input to the binary operator.  

Another important class in R is the `factor`. Factors allow ordering of descriptors/ nominal variables. Many R functions treat factors differently compared to the character class. It is important to be aware if your data is stored as character vs factor.  
The default in R is to store all character data as factor. You can turn this default behavior off with `options(stringsAsFactors = FALSE)`. When we look at the data object class `data.frames` we will have a look at some examples involving factors.  



## data storage objects  
There are several objects in R that handle data storage. We will go through them one by one.   
But first the concept of placeholders, also called identifiers, has to be explained.  

You, as a programmer, are expected to create logical placeholder names. These names should be short, but still clarify what it is. Placeholders are used to store any object type. It is recommended to:  
* separate words with underscores: my_variable.  
* use small letters  
* for functions use verbs: join_tables()  


### placeholders  
As an example of a placeholder, let us define a variable that enables us to store a time-stamp:  
```{r}
current_time <- Sys.time()
current_time  
```
Run the lines several times, see how the placeholder `current_time` stores the output of our `Sys.time()` function?  
We can also store the output of a calculation:  

```{r}
my_result <- 5 * (1+5)
my_result
```
And use this later on in our analysis, because it is stored in `R memory`! 
```{r}
my_result + 1
```

* exercise: explain and then fix the error message you get when running the following line of code: 
```{r}
my_other_result + 1  
```




### vectors
A vector object is created with the `c()` command, each entry is separated by a comma:  
`my_vector <- c( 4, 6, 2, 4, 1)`   


* exercise: Now that we know how to create vectors, how to create variables and strings/ characters lets practice this:  
Create a vector with the `c()` function containing the sentence, "hello, my name is " in the first entry and in the second vector entry a placeholder containing your name.  Make sure to define the placeholder containing your name first, or you will get an error!   
```{r}

```


Another way to create sequences of numbers is with the `seq()` function. Have a look at the help file for `seq' by typing
?seq  
Note the `from` `to` `by` and `length.out` arguments. To create a sequence of 6 equally spaced numbers from 0 to 10 you can do:
`my_seq <- seq(from = 0, to = 10, by = 2)` or you can do `my_seq <- seq(from = 0, to = 10, length.out = 6)` 
Another useful method to generate sequences of integers is the colon: `1:10`  

* exercise: Try all three methods to generate a sequence of numbers from 12 to 2 with steps of 1. 
```{r}

```


It is also possible to do calculations with numeric vectors. There is a behavior to be aware of called recycling. 
* exercise: Calculate the sum of these vectors and explain recycling:  `1:6 + 1:2` also try `1:5` + `1:2`, when and why did you get a warning?   
```{r}

```

Now let us move on to selection operations on vectors. Square brackets `[]` are used for selection. Selection by dimension (rows/columns/ or 3d) can be performed by separating the arguments inside the brackets by comma's. A vector however, has only 1 dimension. To select the second entry of my_seq: `my_seq[2]` 
exercise: Select entry 1, 3 and 5 of `my_seq`:  
```{r}

```

### matrixes  
Single class tables can be stored as matrices. Matrices are not so useful for us biologists, but super useful for fast matrix calculations in the for of linear algebra needed for the various statistical tools available in R.  We will have a quick look at them. To define a matrix:  
```{r}
my_matrix <- matrix(1:9, nrow = 3, ncol = 3)
my_matrix
my_matrix2 <- matrix(1:9, nrow = 3, ncol = 3, byrow = TRUE)
my_matrix2
```
Selection of matrix data is two dimensional, you need row and column identifiers, selection occurs like this: 
`my_matrix[`*row-numbers*`,` *column-numbers*`]` to select row 2 and 3 of the 1st and 3rd columns:  

```{r}
my_matrix[2:3, c(1, 3)]  
```


### data.frames  
The data.frame allows mixing data classes, so you can mix numeric, strings and factors. Let's have a look at the already available `iris` data.frame using the `head` function to display only the first 7 rows.   
```{r}
head(iris, n = 7)
```
To define your own data frame use the `data.frame()` function:  
```{r}
my_dataframe <- data.frame(names = c("Steven", "Margriet", "Gerard", "Elke"),
                           hobbies = c("R", "Python", "cheminformatics", "pharmacokinetics"),
                           sex = c("M", "F", "M", "F"),
                           fakeage = c(38, 30, 34, 28))
my_dataframe
```


The column classes are already displayed under the column names (if you have a recent version of R Studio installed). An alternative to inspect the column classes is to check them manually, for column one and five this would be:  
```{r}
class(iris[, 1])
class(iris[, 5])
```
This mixing of data types is not possible in matrices.  
Before we dive deeper into `data.frames`, we will discuss the `factor` class. Let us print the contents of the `Species` column from the `iris` `data.frame`:  
```{r}
iris[, 5]
```
Note the `Levels` at the bottom? This informs us this column has the class `factor`, and also informs us of the unique factor levels. This is mainly important for statistical analysis, where defining the order is important. However many functions and visualization tools in R also use the factor data type to define for example the order of displaying data labels or the order of plotting nominal values on the x-axis.  
You can create the `factor` class with (first we transform the `Species` to character class): 
```{r}
iris$Species <- as.character(iris$Species)
iris$Species
iris$Species <- factor(iris$Species)
iris$Species
```
 Always think of `data.frames` in a column-wise fashion. A column represents a variable. Next to `iris[, 5]` We can aces `data.frame` columns with `iris$Species` or `iris[, "Species"]`  

This is an opportune moment to explain Boolean and conditional selection. The Boolean class, in R called the logical class, is another class of data. It has the values `TRUE`, `FALSE` and `NA`. The `NA` is used if the entry does not exist.  
With logicals it is possible to perform selection. For example;  
`iris[ iris$Petal.Length < 1.4,  ]` displays all rows of the `iris` data for which the Petal Length is smaller than 1.4.  
printing the `iris$Petal.Length < 1.4` statement clearly shows what is happening.  
```{r}
iris$Petal.Length < 1.4
```
It is possible to perform multiple conditional selections by using the `&` sign in between.  

* exercise: Display the `Petal.Length` for all plants that belong to the virginica Species.  
```{r}

```


* exercise: Display the `Petal.Length` for all plants that belong to the virginica species for which the petal length is smaller than 1.6.    
```{r}

```

A very useful operator for selection to multiple entries is the `%in%` operator:  
```{r}
iris[ iris$Species %in% c("virginica", "setosa")  & iris$Sepal.Length %in% c( 7.2, 5.3),]
```



Finally, we will have a quick look how we can perform calculations with data frames. To define a new variable (thus column) in your data frame you can use direct assignment to a new variable: `iris$is_small_Sepal <- iris$Sepal.Length < 5 | iris$Sepal.Width < 3`   
Now note that you can do calculations on Boolean
```{r}
sum(c(TRUE, TRUE, FALSE, FALSE, TRUE))
```
* exercise: how many small sepal plants are there for each of the three species? With small as defined above.  
```{r}

```

Note how we used multiple columns in a `data.frame` to calculate a new column and that these operations are performed row-wise, thus for each plant!  

### lists  

Lists are another object class in R to store data. Actually any object can be stored in lists, for example functions, data frames, vectors, characters and even other lists can all be placed in a list. If you put 1 object in a list, the list has length 1, putting 2 objects in the list creates a list of length 2, and so on. You can name the entries in the list making access easy and explicit. Let's create a list with named entries:  
```{r}
my_list <- list(teachers = my_dataframe, stat = t.test( 
  subset(iris, Species == "virginica")[, "Sepal.Length"], subset(iris, Species == "setosa")[, "Sepal.Length"]) )

my_list

```

As you can see, the first entry in the list is our `my_dataframe` and the second entry is the result of the `t.test()` function where we tested for different `Sepal.Length` between the species `virginica` and`setosa`  

To access an entry in our list there are several options. One can use the name:  
```{r}
my_list$teachers
my_list$stat
```
Or simply index numbers:  
```{r}
my_list[[1]]
```
If you want to keep the list attributes use a single bracket:  
```{r}
my_list[1]
```

To select multiple slots simultaneously use single brackets:  
```{r}
my_list[1:2]
```

Accessing objects inside lists is also straightforward, access the list entry and then treat the object selections as you would normally:  
```{r}
my_list$teachers$hobbies
```

Or you could do:  
```{r}
my_list[[1]][, "hobbies"]
```

* exercise: Store the iris data frame as and additional slot in the existing list object `my_list`  
```{r}

```

* exercise: Display the `Sepal.Lengths` of all `virginca` and `setosa` plants from within the `my_list`  
```{r}

```



To conclude, lists are a useful addition to data frames because they are more flexible.  


### arrays  
Arrays are higher dimensional  matrices. For example, a three dimensional array can be thought of as multiple matrices stacked behind one another, for 4 3x3 matrices:    
```{r}
my_array <- array(1:28, dim = c(3, 3, 4))
dim(my_array)
my_array
```

To access 2nd row, 2nd column of 2nd matrix: [row, column, extra_dimension1]  
```{r}
my_array[ 2, 2, 2 ]
```
Or to access the 3rd matrix:  
```{r}
my_array[ , , 3]
```

But you could also give the matrices themselves indexes in multiple dimensions, for example:  
```{r}
my_array <- array(1:28, dim = c(3, 3, 2, 2 ))
dim(my_array)
my_array
```



## control flow and repetition  
Next to one time diving into data for analysis, it is often useful to create automated workflows, your own packages or even pieces of software. In these cases it is useful to learn a little bit more than loading and accessing data, so we will dive into a bit more advanced programming tools, such as control flow and repetition. These concepts are often also useful even for interactive data analysis.  

### conditionals

Conditionals make your program perform an action depending on the input. To do this the `if else` syntax is used. 
The basic syntax is:  `if(TRUE) { do this }`  
For example;  
```{r}
sepal_size <- 6
if( sepal_size < 5 ) { # note the opening curly braces  
  sepal_class <- "small"
} # note the matching closing curly braces  
```

* exercise: run the lines above, what is sepal_class now? Does it exist? Why (not)?  

So the `if` statement above was evaluated as `FALSE` and the lines between the curly braces were skipped. Wouldn't it be nice to define sepal_class as `large` without creating another separate `if` statement? Rejoice, we can!  

```{r}
sepal_size <- 6
if( sepal_size < 5 ) { 
  sepal_class <- "small"
} else {
  sepal_class <- "large"
}
```


It is even possible to create as many `else if` conditionals as you like:  
```{r}
sepal_size <- 6
if( sepal_size < 5 ) { 
  sepal_class <- "small"
} else if( sepal_size > 8) {
  sepal_class <- "large"
} else {
  sepal_class <- "medium"
}
```
* exercise: without running the code above, what does `sepal_class` evaluate to?  

* exercise: modify the example above to include another interval between 5 and 6 that evaluates to `medium-small`  

### for loops
Coding would be a lot less useful if we would not be able to repeat operations easily. For repetition R uses the following loop syntax: `for(i in 1:10)`{ code to repeat in here }.  The `i` can be any made up placeholder name, and the number of repetitions can be to your choosing as well. In this example we perform 10 repetitions with the sequence of numbers from 1 to 10. Take note the variable `i` will evaluate to each of these numbers in the sequence. This is useful for indexing as will become clear in the following example:  
```{r}
for(i in 1:5) {
  print(paste0("is is: ", i))
  result <- i + 1
  print(paste0("and i + 1 is: ", result))
  }

```

* exercise: create a for loop like in the example above, but let `i` evaluate to the sequence 1, 3, 5, 7, 9 instead.  
```{r}

```


* exercise: modify the script you just created so that it only prints values that are larger than 4 and smaller than 8. You will need to add an `if` statement inside the loop somewhere!   
```{r}

```

* exercise: add a third gender to the script below:  
```{r}
gender <- readline(prompt = "What is your gender, M/F?")
if(gender == "M"){
  print("Hi Sir")
} else if(gender == "F") {
  print("Hi Mam")
} else {
  print("Please input M or F")
}

```

* final exam question: create a small gambling game. Here are the rules:  
** you start with 5 euro.  
** you loose when you have 0 or less euro.  
** you have to guess how the dice will fall. 
** if you guess correct you win 3 euro  
** if you guess incorrect you loose 1 euro  
** use a loop to play 5 times, you will ask for user input 5 times  
** at each iteration inform the player how much money is left    
Hints:
This is how you can program a dice in R: `sample(1:6, 1)`
You will need the `readline` for user input (see previous exercise)
You can stop a loop with `break`  

```{r}
my_money <- 5
for( i in 1:5) {
wn <- readline(prompt = "what number")
rd <- sample(1:6, 1)
if( wn == rd) {
  my_money <- my_money + 3
} else{
  my_money <- my_money - 1
}
if(my_money < 1) {
  print("you loose")
  break
}
print(paste0("you have: ", my_money))
}
```

There is a way to make the game better, by using a `while` loop. Here is an example:
```{r}
while(my_money > 0 | my_money < 10 ) {
  
}
```

* exercise: modify your game by replacing the `for` loop with a `while` loop.  
```{r}

```



## functions 

We have come across many functions already. It is good to be aware of the origin of functions. R has a system to look for functions in different environments, this is to ensure identical function names from different packages do not interfere with one another.  
The base R functions, like `mean`, `sample`, `print`  etc come with every installation of R.  
Functions created by package developers become available after loading a package.  
User defined packages are created by the user during a R session. 
To define your own function use the following syntax:  
```{r}

my_function <- function(input_arguments) {
# code that does stuf with input_arguments and returns something
  return(output)
}
my_function(input_arguments = 5)

```

A working function:  
```{r}
sum_two_numbers <- function(a, b) {
  output <- a + b
  return(output)
}

# After the function is defined you can use the function:  
sum_two_numbers(a = 5, b = 6)
```

It is good practice to be explicit, however this also works:  
```{r}
sum_two_numbers <- function(a, b) {
  a + b
  }

sum_two_numbers(5, 6)

```


We will have to delve into environments first before we can continue with functions  

### packages & environments & namespaces 

In R it is relatively straightforward to create you own package. If accepted by the maintainers of CRAN, the repository for R packages, it will become available for anyone by a simple call to the `install.packages()` function! But you can also put your package or scripts on Git Hub, and make it available that way.  
With a simple call to the explicit function name (no parenthesis), for example `mean` you can see the `environment: namespace:base` - which means the function is available after installing R.  
Now try to make a call to `trapz`.  
```{r}

```

Indeed, "object `trapz` not found".  
If you do not want to code a trapezium based function to calculate the area under a curve yourself, you can Google or search CRAN for a package that someone already wrote to do this. For example the `pracma`  package will do this for you. Install the package:
```{r}
#install.packages("pracma")
```
After installing the package, the package will be located in your local `library` folder. Which you can find by typing `.libPaths()`  

```{r}
.libPaths()
dir(.libPaths()) # the dir function lists the directories within the path specified  
```
So you only have to install a package once and then R can find it. However if you want to use it during an R-session you have to load it:  
```{r}
library(pracma)
```

Now we type `trapz`:  
```{r}
trapz
```

You get the source code and at the end you see the function is from the pracma namespace. One final remark on environments: When you try to evaluate a function (or any object for that matter) R will search through the environments in a certain order and use the first it finds. To find out where and in what order R will search you can use:  

```{r}
searchpaths()
```
However it is important to realize if a call is made within a function, this environment is searched first! Thus, also functions have there own environment, which only exists inside the function where the call is made. Here is an example of how this works, make sure to understand this as it is important:  
```{r}
a <- 1

my_fun <- function(input) {
  a <- 5
  output <- a * input
return(output)
  }

my_fun(input = 1)


my_fun2 <- function(input) {
  
  output <- a * input
return(output)
  }

my_fun2(input = 1)


```
So `my_fun` finds `a` in it's own environment, but `my_fun2` does not so it searches outside the function environment. 
R is an interactive programming language, the danger with interactive programming is that you might have used certain variable names before and forgot about it. Which can easily happen during extensive data analysis projects. This is why it is good practice not to clutter your R-memory with variables. Keep it clean by using functions. Observe how defining a variable in a function is not stored in global R memory:

```{r}
my_fun3 <- function( b0, b1, b2, x) {
  y <-  b1*x + b2*x
  output = y + b1
return(y)
  }

my_fun3(b0 = 2, b1 = 0.1, b2 = 0.5, x = 1:10)

b0
b1
b2
x
y
output

```
Notice how none of the variables exist! If you did not use a function, R memory would be cluttered already.  
There are more reasons functions are useful, they are easy to reuse, and needed for functional programming, which we go into in the next section.  

By the way, you can see your R memory and the objects that live inside in the top right panel in `Global Environment`. You can also type `ls()`  

* exercise: create a function with input arguments your name, age and weight. In the function body calculate your BMI `weight in kg`/ `(height in cm)^2`. Store your name, age, weight and BMI in a named list and return this list.  
```{r}

```


### functional programming: apply family  
Functional programming: "Functions that do stuff or use other functions" is close enough for us.  
Two important functions to this end are `lapply` and `sapply`. The difference is only in how they output the result. `lapply` outputs a list, and `sapply` tries to format the output - which often wont' be satisfactory - so try `sapply` first, then `lapply`.  
To quickly check the classes of the columns of a data frame:
```{r}
sapply(my_dataframe, class)
```
As you can see, the arguments are: first the object to perform a function over, then the function that is called per sub index (in this case the columns) in the object. For `lists`:   
```{r}
sapply(my_list, class)
```

You can also use sapply instead of loops:
```{r}
sapply(1:10, function(i) { i * 2})
```
This might be a tricky one, but what happens is that sapply subsets the columns , list entries or entries in a vector, and feeds this 1 by 1 to the function. In the last example we simply defined our own function, and gave the fed-subsets a name: `i` - which we can then continue to use in our function.  

* exercise: Define a function that adds a value to each subset of the vector 1:10. Use `sapply` to do this. The value that you add should be an argument to the function    
```{r}

```

* exercise: calculate the summary of each `iris` column using sapply or lapply using the `summary()` function  
```{r}

```

Hopefully you noticed that if you want to specify some arguments to the function within an apply type of function, you need to explicitly name the subsets fed into the function by the apply function. You can do this by writing the whole function: `function(x) { function code or existing function name which used the x argument which contains the fed subset }` 



## dplyr & tidyr 

The remainder of the course will be about data formatting, summarization and visualization. It is entirely possible to do these things using base R. However in recent years better alternatives have been developed. Now that the basic R syntax has been explained, it is time to introduce a more consistent and easier to read syntax created by Hadley Wickham in the form of key packages `dplyr`, `tidyr` and `ggplot2`. Most operations can be performed with `dplyr` and `tidyr`, and `ggplot2` can handle most visualizations. It is however still necessary to know the basic R syntax - especially for more complicated and specific statistical analysis for which many methods exist. We will explain method (dispatch) in the visualization paragraph.         

First install and load the packages:  
```{r}
#install.packages(c("dplyr", "tidyr", "ggplot2"))
library(dplyr)
library(tidyr)
library(ggplot2)

```

### tidyr & dplyr syntax

Advantages:  
** Piping `%>%` more readable: data flow operator from left to right    
** Tibble, nicer implementation of data.frame  
** Consistent language: `filter`, `select`, `mutate`,  `summarise` and `arrange`  
** Explicit function names (function names do what they say)  

Overview:  
row selection:  `filter(iris, Sepal.Length >6)`  
column selection: `select(iris, Species : Petal.Length)`  
Calculations: `mutate(iris, Petal.Ratio = Petal.Width / Petal.Length)`  
Define annotation: `group_by(iris, Species)` For summarizing or joining tables  
Summarize based on group annotation: `summarise(iris, n = n(), mean.Petal.Ratio = mean(Petal.Ratio))` 
Row order the table: `arrange(iris, desc(mean.Petal.Ratio))` 
Merging tables based on a shared column id: `left_join()`, `full_join()` 

For example, using the dplyr syntax on `my_dataframe`:  
```{r}
my_dataframe %>% filter( names %in% c("Margriet", "Elke") ) %>%
  select(names:hobbies) # select columns from names to hobbies
```
To perform calculations on a data frame and perform a statistical summary:  
```{r}
my_dataframe %>% 
  mutate(nonsensicle_confidence_limits = 1.97*sd(fakeage)/sqrt(3)) %>%
  group_by(sex) %>%
  summarize(n = n(), total_age = sum(fakeage), CoV = sd(fakeage)/total_age )
    
```
In the example above the `mutate` call calculates the confidence limits and adds it as a new column.  
The `group_by` call is needed to group the data frame by sex, so that R knows over what levels to summarize over in the next line. The `n = n()` gives us the number of rows summarized over, then two more calculations are performed as extra arguments to the `summarise` call; the sum of the age of the males/females, and the coefficient of variation.  


* exercise: After observing the following code, rewrite it with pipes and `dplyr` syntax: 
hint: run line by line and observe the result, some functions like `aggregate` have not been treated but by looking at the result will make its function it obvious.  
```{r}
tmp <- iris[ iris$Sepal.Length > 6, ]
tmp <- tmp[, c("Species", "Petal.Width", "Petal.Length")]
tmp$Petal.Ratio <- tmp$Petal.Width / tmp$Petal.Length
tmp2 <- aggregate(Petal.Ratio ~ Species, data = tmp, FUN = length)
tmp <- aggregate(Petal.Ratio ~ Species, data = tmp, FUN = mean)
colnames(tmp)[2] <- "mean.Petal.Ratio"
colnames(tmp2)[2] <- "n"
tmp <- merge(tmp2, tmp, by = "Species")
tmp[ rev(order(tmp$mean.Petal.Ratio)), ]
```

```{r}

```

## data reshaping, summarization and merging  

Data reshaping, summarization and merging are fundamental data analysis operations. There are a multitude of ways this can be done in R, this can make R very confusing. For this reason we will stick to the `dplyr/tidyr` syntax.   
To explain these concepts and to demonstrate use-cases, we will perform a small case study.  
Carefully run the lines and make sure to understand each part.  

Observe the `iris` data set:
```{r}
head(iris)
```

Our goal is to calculate the average sepal and petal lengths and widths and add it to the existing iris table. Let us do this now, we first define the groups, then calculate the mean and finally add the result of this calculation to the iris table:    

```{r}

mean_iris <- iris %>% group_by(Species) %>%
  summarise_all(mean) 
  
iris <- left_join(iris, mean_iris, by = "Species", suffix = c( "", ".mean") )
head(iris)

```
So for each measure- column we have to create an extra column to store the mean. A better way would be to first `gather` the measure-columns into a single column:  

```{r}
# first reload iris:
data(iris)
head(iris)
iris %>% gather(key = variable, value = value , -Species)

```
Basically: `gather(data, key = "new name of combined columns", value = "new name of value column", -name of columns not to gather in key column)`
Or you could specify the columns to gather: 
```{r}
iris %>% gather(key = variable, value = value,
       Sepal.Length, Sepal.Width, Petal.Length, Petal.Width)
```

Now let's use what we have learned to streamline our case study objective:  
```{r}
iris %>% gather(key = variable, value = value , -Species) %>% 
  group_by(Species) %>% 
  mutate(mean_value = mean(value))
```
Then there is also `spread`, which creates the wider format of data:  

```{r}
iris <- iris %>% mutate(id = 1:nrow(iris)) # add id so we can go from long format back to wide format
iris_long <- iris %>% gather(key = variable, value = value , -(Species:id)) 
iris_wide <- iris_long %>% spread( key = variable, value = value )
```
basically: `spread(longdata, key = name of column to spread, name of value column)` 


* exercise: start with the iris data, go to long format with `gather` then back to the original format, but now without using the `id` column. You can reload/ clean the iris data with `data(iris)`  Did this work? Why (not)? What does the `id` represent?  
```{r}

```


* exercise: Create a table that contains the max value of each measurement for each species from the `iris` data. What species has the plant with the largest petals and sepals?  
```{r}

```


## data visualization  
Most useRs tend to use the `ggplot2` package for data visualization. However before we delve into `ggplot2` we will explain the base r plotting system, because many useful methods exists for this system, especially in the context of statistical analysis like residual plots.  

### base plotting methods  
With base plotting you can modify the graphs in any little details you would like, this is however not very intuitive and requires quite a large amount of time to master. It is easier and quicker to do with the `theme_` system from the `ggplot2` package. In addition we can only recommend to modify little details using vector based images and import these into for example illustrator (since the university has an faculty wide license anyway).    




### method dispatch  
Method dispatch is everywhere in R. Functions call methods depending on the class of the input argument to the function. You can display the available methods for the plot function with `methods(plot)`. We will show you an example using the `plot` function, first we create three different object classes, then observe the behavior of the plot function:      

```{r}
dist/ hclust
lm()
data.frame
```

Also the type of data will change the way `plot` behaves. Compare for example nominal vs numerical data:  

```{r}

```





### ggplot2   

melt case  



## read and write functions   





